Statically Typed String Sanitation Inside a Python

### 2014-10-07 sanitation-psp14/sanitation-psp14-short.tex
Updated through 2.2

\begin{lem}[Properties of Regular Languages.] \label{thm:regexprops}
~
\begin{enumerate}
\item If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1r_2}$.
\item For all strings $s$ and regular expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
\item Regular languages are closed under reversal.
\end{enumerate}
\end{lem}

@@

\begin{lemma}[Canonical Forms for String Fragment of $\lambdas$]\label{thm:cfs}
  If $\emptyset \vdash e : \stringin{r}$ and $e \sreduces v$ then $v=\rstr{s}$. 
\end{lemma}
\begin{proof}
The proof is again by induction on the typing relation, following straightforwardly.
\end{proof}

@@

<$\iota$> ::= $x$ | $\lambda x . \iota$ | $\iota(\iota)$ \hfill target terms \alt
$\tstr{s}$ | $\tconcat{\iota}{\iota}$ | $\pstrcase{\iota}{\iota}{x,y.\iota}$ \alt
  $\rx{r}$ | $\preplace{\iota}{\iota}{\iota}$ | $\tcheck{\iota}{\iota}{\iota}{\iota}$

  <$\dot{v}$> ::= $\lambda x . \iota$ | \tstr{s} | \rx{r} \hfill target values

@@

The system $\lambdap$ is another straight-forward extension of a simply typed lambda calculus
with a $\str$ type and a $\regex$ type, as well as 
some operations -- such as concatenation and replacement -- found in the standard libraries of many programming languages.
The operations of $\lambdap $ correspond to ``run-time'' versions of the operations performed statically in $\lambdas$ in a way made precise by the translation rules described in the next section.
%Unlike $\lambdas$, $\lambdap$ does not statically track the effects of string operations.
The language $\lambdap$ is so-called because it is reminscent of popular web programming languages,
such as \textbf{P}ython or \textbf{P}HP. We will discuss an implementation within the former in the next section.

The grammar of $\lambdap$ is defined in Figure~\ref{fig:lcsSyntax}.
The typing rules \textsc{P-T-} are defined in Figure~\ref{fig:slambdap}
and a big-step semantics is defined by the rules \textsc{P-E-} in Figure~\ref{fig:dlambdap}.

@@

Type safety for $\lambdap$ is straight-forward, but is necessary in order to establish the 
correctness of our translation.
Again, we elide a more careful treatment by treating a special case where our language terminates but
note the multiple approaches to proving soundness for non-terminating languages with natural semantics \cite{coinductionoo, stepindexing, coinduction}.

\begin{lem}[Well-formedness of regular expressions]
If $\tctx \vdash \iota : \regex$ then $\iota \treduces \rx{r}$ such that $r$ is a well-formed regular expression. 
\end{lem}

\begin{thm}[Safety for $\lambdap$]
  Let $\iota$ be a term in the target language. If $\tctx \vdash \iota : \tau$ 
  then $\iota \sreduces \iota'$ and $\tctx \vdash \iota' : \tau$.
\end{thm}

@@

The translation from $\lambdas$ to $\lambdap$ is defined in figure 5.
The coercion cases are most interesting. If the safety of coercion in manifest in the
types of the expressions, then no runtime check is inserted.
If the safety of coercion is not manifest in the types, then a check is inserted.

>>>>

\begin{lem}[Properties of Regular Languages.] \label{thm:regexprops}
~
\begin{enumerate}
\item If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1\cdot r_2}$.
\item For all strings $s$ and regular expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
\item Regular languages are closed under reversal.
\end{enumerate}
\end{lem}

@@

\begin{lem}[Canonical Forms for String Fragment of $\lambdas$]\label{thm:cfs}
  If $\emptyset \vdash e : \stringin{r}$ and $e \sreduces v$ then $v=\rstr{s}$. 
\end{lem}
\begin{proof}
The proof is again by induction on the typing relation, following straightforwardly.
\end{proof}

@@

<$\iota$> ::= $x$ | $\lambda x . \iota$ | $\iota(\iota)$ \hfill target terms \alt
$\tstr{s}$ | $\tconcat{\iota}{\iota}$ | $\pstrcase{\iota}{\iota}{x,y.\iota}$ \alt
  $\rx{r}$ | $\tcheck{\iota}{\iota}{\iota}{\iota}$ | $\preplace{\iota}{\iota}{\iota}$

  <$\dot{v}$> ::= $\lambda x . \iota$ | $\tstr{s}$ | $\rx{r}$ \hfill target values

@@

The system $\lambdap$ is another straight-forward extension of a simply typed lambda calculus
with a $\str$ type and a $\regex$ type, as well as 
some operations -- such as concatenation and replacement -- found in the standard libraries of many programming languages.
The operations of $\lambdap $ correspond to ``run-time'' versions of the operations performed statically in $\lambdas$ in a way made precise by the translation rules described in the next section. 
%Unlike $\lambdas$, $\lambdap$ does not statically track the effects of string operations.
The language $\lambdap$ is so-called because it is reminscent of popular web programming languages,
such as \textbf{P}ython or \textbf{P}HP, albeit statically typed. We will discuss an implementation within the former in the next section.

The grammar of $\lambdap$ is defined in Figure~\ref{fig:lcsSyntax}.
The typing rules \textsc{P-T-} are defined in Figure~\ref{fig:slambdap}
and a big-step semantics is defined by the rules \textsc{P-E-} in Figure~\ref{fig:dlambdap}. The semantics are straightforward given the discussion of the corresponding operations in the previous section.

@@

Type safety for $\lambdap$ is straight-forward, but is necessary in order to establish the 
correctness of our translation.
Again, we elide a more careful treatment by treating a special case where our language terminates per the discussion above.

\begin{thm}[Safety for $\lambdap$] If $\emptyset \vdash \iota : \tau$ 
  then $\iota \sreduces \dot v$ and $\tctx \vdash \dot v : \tau$.
\end{thm}

We can also define canonical forms for regular expressions and strings in the usual way:

\begin{lem}[Canonical Forms for Target Language]
~
\begin{enumerate}
\item If $\emptyset \vdash \iota : \regex$ then $\iota \treduces \rx{r}$ such that $r$ is a well-formed regular expression. 
\item If $\emptyset \vdash \iota : \str$ then $\iota \treduces \tstr{s}$.
\end{enumerate}
\end{lem}

@@

The translation from $\lambdas$ to $\lambdap$ is defined in Figure \ref{fig:tr}.
The coercion cases are most interesting. If the safety of coercion in manifest in the
types of the expressions, then no runtime check is inserted.
If the safety of coercion is not manifest in the types, then a check is inserted.
